prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-29 inst/MyArticle/paper.Rmd
text updates

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r Biocpkg("limma")` [@Ritchie2015], `r Biocpkg("MSstats")` [@Choi2014], `r Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r Biocpkg("msqrob2")` [@Goeminne2016] or `r Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, `r Biocpkg("MSstats")` determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r Biocpkg("limma")` allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r Biocpkg("PECA")` performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from `r Biocpkg("limma")` or `r Biocpkg("PECA")`, to protein level estimates. Furthermore, `r Biocpkg("msqrob2")` combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The `r Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, `r Biocpkg("MSstats")` handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the R package `r Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

We can use all the R packages discussed when analyzing parallel-group designs using a single explanatory variable and contrasting groups. However, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that, for instance, `r Biocpkg("limma")` and `r Biocpkg("proDA")` allow us to fit a comprehensive variety of models and test various hypotheses; however, good knowledge of the design matrix specification using the R formula interface is required [@law2020guide].

@@

We use the Ionstar [@shen2018ionstar] and CPTAC (available from `r Biocpkg("msdata")`) dataset to benchmark the modeling methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r Biocpkg("MSstats")` and `r Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$).

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an MSFragger[@yu2020fast] configuration can be found in the Appendix. We implemented methods that transform the data into tidy data for popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (see Supplementary Figure SX).

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r Biocpkg("proDA")` or of SAINTexpress [@teo2014saintexpress] used to analyze data from protein interaction studies.

(ref:ContrastUML)  Unified modeling language (UML) diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ (Table \@ref(tab:prolfquaModels)), reference methods to fit models, and compute contrasts. The model builder method fits the statistical model given the data and a strategy. The models are used to analyze variance (ANOVA) or to estimate contrasts. All classes estimating contrasts implement the _ContrastsInterface_. Results of external tools, e.g., SAINTexpress, or proDA are adapted to implement the Contrasts Interface.

@@

@@

To evaluate the performance of differential expression analysis, we use the IonStar benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$, $4.5\%$, $6\%$, $7.5\%$ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_ and _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, hence 20 raw files in total. We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata]. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

@@

@@

@@

The peptide abundances (from the MaxQuant _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r Biocpkg("PECA")` model, and hurdle model `r Biocpkg("msqrob2")` to peptide intensities.

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which MaxQuant reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`'s functionality (see Supplementary Figure S[X]). 
To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric,@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula and which comparisons to compute from the sample annotation provided in tabular form, similarly to SAINTexpress and `r Biocpkg("MSstats")`. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with the results produced by `r Biocpkg("MSstats")`, `r Biocpkg("proDA")` and `r Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4178 proteins with at least two peptides, there are in total $16712$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r Biocpkg("MSstats")` estimates $16038$ group differences while the mixed effect models estimates the fewest with $15756$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, which indicates that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

@@

Using a benchmark dataset whose ground truth is known (see Methods), we assessed the performance of different modeling approaches implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")`, `r Biocpkg("MSstats")`, `r Biocpkg("proDA")` and `r Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes which methods we have evaluated, which MaxQuant files we used as input, and if the models are fitted to peptide or protein intensities. We make the R-markdown files to replicate the benchmarking available at `r BiocStyle::Githubpkg("wolski/prolfquabenchmark")`.

@@

The IonStar/MaxQuant dataset (see Methods) captures only the variance from the chromatography, electro-spray, and mass spectrometric measurements since only technical replicates are available for each dilution. Therefore essential sources of variation typically present in other experiments, such as the biochemical and biological ones, are not measured. Furthermore, this dataset with a parallel group design does not allow for benchmarking models with interactions. However, other benchmark datasets have similar constraints [@edwards2015cptac, @o2018proteome]. Furthermore, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a complete factorial design with an up-to-date version of MaxQuant. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret our findings. Specifically, the observed variances will be higher in datasets with biological replicates, and the power will be lower for the same number of samples. Furthermore, the proportion of missing observations in real-life datasets might be higher or distributed differently in groups.

@@

The R-package `r Biocpkg("proDA")`, `r Biocpkg("msqrob2")` and `r Biocpkg("prolfqua")` do not impute missing data but integrate them into the statistical model, while `r Biocpkg("MSstats")` filters and imputes the data using an accelerated failure model. Despite imputation, `r Biocpkg("MSstats")` estimates group differences for less proteins, and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:figure3) Panel C shows that when using `r Biocpkg("MSstats")` the proportion of false discoveries might be very high for a low $FDR$ because of false positives. Of note, MSStats uses the _evidence.txt_ file, while all the other methods use _peptide.txt_ files, as input (see Table \@ref(tab:describeBenchmarkedModels))), a possible confounding factor to consider, but we can not control for.

We obtained difference and FDR estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (*proDA*), b) the hurdle model (*msqrob2*), and c) `prolfqua_merge` (Table \@ref{tab:describeBenchmarkedModels}). We observe that the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is a significant difference between the $pAUC_{10}$ for all three methods, but did not reject the null hypothesis that there is no such difference (Supplementary Section "Comparing pAUC for IonStar/MaxQuant dataset"). Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods. Furthermore, all three models perform similarly well when examined using CPTAC/MaxQuant dataset (see Supplementary section "Benchmarking msqrob2, proDA and prolfqua using the CPTAC dataset"). In addition we also examined the performance of the *proDA* and `prolfqua_merged` model using the IonStar dataset but processed with the `FragPipe v15` software (see Supplement). While, for the IonStar/MaxQuant dataset, *msqrob2* performed best, *proDA* performed best for the CPTAC/MaxQuant data, and `prolfqua_merged` for the IonStar/FragPipe dataset.

>>>>

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r Biocpkg("limma")` [@Ritchie2015], `r Biocpkg("MSstats")` [@Choi2014], `r Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r Biocpkg("msqrob2")` [@Goeminne2016] or `r Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, *MSstats* determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, *limma* allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package *PECA* performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from *limma* or *PECA*, to protein level estimates. Furthermore, *msqrob2* combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The *proDA* package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, *MSstats* handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the R package *proDA* models missingness using probabilistic dropout models [@bioxrvproDA2020].

We can use all the R packages discussed when analyzing parallel-group designs using a single explanatory variable and contrasting groups. However, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that, for instance, *limma* and *proDA* allow us to fit a comprehensive variety of models and test various hypotheses; however, good knowledge of the design matrix specification using the R formula interface is required [@law2020guide].

@@

Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$). We use the IonStar [@shen2018ionstar] and CPTAC[@edwards2015cptac]  datasets, processed with *MaxQuant* and *FragPipe* to benchmark the modeling methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r Biocpkg("MSstats")`, `r Biocpkg("msqrob2")` and `r Biocpkg("proDA")`.

Although other benchmark datasets exist[@wessels2012comprehensive;@o2018proteome], the IonStar dataset has the advantage that the expected differences of the spike in proteins among groups are slight compared to other benchmark datasets, making DEA more difficult and enabling us to see performance differences among the methods. For the same reason of making the DEA more challenging and enhancing the differences between the statistical modeling approaches, we focus on the MaxQuant outputs since they are noisier than those produced by FragPipe.

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an MSFragger[@yu2020fast] configuration can be found in the Appendix. We implemented methods that transform the data into tidy data for popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supplementary Figure SX) in RStudio [@RStudio2022], and help to implement argument free functions.

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r Biocpkg("proDA")` or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

(ref:ContrastUML)  Unified modeling language (UML) diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ (Table \@ref(tab:prolfquaModels)), reference methods to fit models, and compute contrasts. The model builder method fits the statistical model given the data and a strategy. The models are used to analyze variance (ANOVA) or to estimate contrasts. All classes estimating contrasts implement the _ContrastsInterface_. Results of external tools, e.g., *SAINTexpress*, or *proDA* are adapted to implement the Contrasts Interface.

@@

### IonStar

To evaluate the performance of differential expression analysis, we use the IonStar benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$, $4.5\%$, $6\%$, $7.5\%$ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_ and _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, hence 20 raw files in total. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

However, other  have similar constraints. Furthermore, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a complete factorial design with an up-to-date version of MaxQuant.

@@

We processed the raw data of the IonStar dataset using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings for orbitrap data. The text files generated by MaxQuant are available in the *prolfquadata* R package [@prolfquadata]. MaxQuant produces various output files which can be used for DEA. We are using the quantification results reported in the 'peptide.txt' file for DEA. However, *MSstats* is using the 'evidence.txt' file for the DEA.

@@

We processed the raw data of the IonStar dataset using the FragPipe [@yu2020fast] Version 14, with default workflow for label free quantification with match between runs. The text files generated by FragPipe are available in the *prolfquadata* R package [@prolfquadata].
Similarily to MaxQuant, the FragPipe produces various outputs which can e used for DEA. We used the total protein intensities reported in the `combined_protein.tsv` file as input for the DEA and call this dataset _IonStar/FragPipe/combined_protein.tsv_ Alternatively, we benchmarked the DEA using the MSstats.tsv file as input and call this dataset _IonStar/FragPipe/combined_protein.tsv_.

@@

We used the CPTAC dataset, available in the R package *msdata*, and described in [@Goeminne2016]. In brief, the Sigma Universal Protein Standard mixture 1 (UPS1) containing $48$ different human proteins was spiked in a protein background of 60 ng/$\mu L$ Saccharomyces cerevisiae strain BY4741. Two different spike-in concentrations were used, 6A ($0.25$ fmol UPS1 proteins/$\mu L$) and 6B ($0.74$ fmol UPS1 proteins/$\mu L$). Three replicates are available for each concentrations. The data were searched with *MaxQuant* version `1.5.2.8.`.

@@

The peptide abundances (from the *MaxQuant* _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r Biocpkg("PECA")` model, and hurdle model `r Biocpkg("msqrob2")` to peptide intensities.

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which *MaxQuant* reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`'s functionality (see Supplementary Figure S[X]).

To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric;@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula, and which comparisons to compute from the sample annotation provided in tabular form, similarly to *SAINTexpress* or *MSstats*. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with the results produced by *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4178 proteins with at least two peptides, there are in total $16712$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that *MSstats* estimates $16038$ group differences while the mixed effect models estimates the fewest with $15756$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, which indicates that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of *MSstats*, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

@@

Using a benchmark dataset whose ground truth is known (see Methods), we assessed the performance of different modeling approaches implemented in *prolfqua*, *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes which methods we have evaluated, which *MaxQuant* files we used as input, and if the models are fitted to peptide or protein intensities. We make the R-markdown files to replicate the benchmarking available at `r BiocStyle::Githubpkg("wolski/prolfquabenchmark")`.

@@

The IonStar/MaxQuant dataset (see Methods) captures only the variance from the chromatography, electro-spray, and mass spectrometric measurements since only technical replicates are available for each dilution. Therefore essential sources of variation typically present in other experiments, such as the biochemical and biological ones, are not measured. Furthermore, this dataset with a parallel group design does not allow for benchmarking models with interactions. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret our findings. Specifically, the observed variances will be higher in datasets with biological replicates, and the power will be lower for the same number of samples. Furthermore, the proportion of missing observations in real-life datasets might be higher or distributed differently in groups.

@@

The R-package *proDA*, *msqrob2* and *prolfqua* do not impute missing data but integrate them into the statistical model, while *MSstats* filters and imputes the data using an accelerated failure model. Despite imputation, *MSstats* estimates group differences for less proteins, and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:figure3) Panel C shows that when using *MSstats* the proportion of false discoveries might be very high for a low $FDR$ because of false positives. Of note, *MSStats* uses the _evidence.txt_ file, while all the other methods use _peptide.txt_ files, as input (see Table \@ref(tab:describeBenchmarkedModels))). Furthermore, *MSstats* uses equalized medians normalization, while all the other methods use robust scaling (see Methods). These are possible confounding factors to consider. While *prolfqua*, as well as *proDA*, is highly modular, and to a lesser extent *msqrob2*, enabling us to use the same data preprocessing and normalization, *MSstats* is a monolithic software, making it unfeasible to change the preprocessing or normalization. Therefore, we could not control these confounders.

We obtained difference and FDR estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (*proDA*), b) the hurdle model (*msqrob2*), and c) `prolfqua_merge` (Table \@ref{tab:describeBenchmarkedModels}). We observe that the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is a significant difference between the $pAUC_{10}$ for all three methods, but did not reject the null hypothesis that there is no such difference (Supplementary Section "Comparing pAUC for IonStar/MaxQuant dataset"). Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods. Furthermore, all three models perform similarly well when examined using CPTAC/MaxQuant dataset (see Supplementary section "Benchmarking msqrob2, proDA and prolfqua using the CPTAC dataset"). In addition we also examined the performance of the *proDA* and `prolfqua_merged` model using the IonStar dataset but processed with the `FragPipe v14` software. While, for the IonStar/MaxQuant dataset, *msqrob2* performed best, *proDA* performed best for the CPTAC/MaxQuant data, and `prolfqua_merged` for the IonStar/FragPipe dataset.

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplement we show the benchmarking results for *proDA* and *prolfqua* when the IonStar dataset is prepossessed with the FragPipe v14 [@yu2020fast] quantification pipeline. Although the number of proteins identified with two peptides is slightly lower with FragPipe than with MaxQuant (3929 vs. 4046), the $pAUC_{10}$ is significantly higher ($81%$ vs. $72%$).
